\chapter{Skills}
\chaplabel{skills}

The success of a wizard is not determined simply by the spells they know and the magics they can perform.
Equally important is their skill at wielding that magic, and even their skill in the more mundane aspects of life.
It's one thing to be able to throw fire, but without the ability to hit what you're aiming for, you're going to be burning down a lot of buildings you didn't mean to.
Similarly, while magic can let you take on someone else's face and voice, doing so is pretty useless if you're a terrible actor.

As such, classes don't simply teach students new techniques; they always contain a practical component to allow students to practise their skills.
Furthermore, schools encourage students to participate in extracurricular activities outside class hours, to help them develop into well-rounded individuals.
All in all, there are many skills, both magical and mundane, that students can develop throughout their education.
These are listed in the following table, while details of each skill are provided in the following sections.

\simpletable{XX}{
	\toprule
	\capital{\mundaneskillbare} & \capital{\magicskillbare}\\
	\midrule
	\capital{\skillref{aim}} & \capital{\skillref{divination}}\\
	\capital{\skillref{athletics}} & \capital{\skillref{elementalism}}\\
	\capital{\skillref{beast-handling}} & \capital{\skillref{enchanting}}\\
	\capital{\skillref{charm}} & \capital{\skillref{illusion}}\\
	\capital{\skillrefspeciality{craft}{\variousspeciality}} & \capital{\skillref{kinetics}}\\
	\capital{\skillref{fight}} & \capital{\skillref{mentalism}}\\
	\capital{\skillref{investigation}} & \capital{\skillref{metamagic}}\\
	\capital{\skillref{leadership}} & \capital{\skillref{necromancy}}\\
	\capital{\skillrefspeciality{lore}{\variousspeciality}} & \capital{\skillref{restoration}}\\
	\capital{\skillref{observe}} & \capital{\skillref{transmutation}}\\
	\capital{\skillrefspeciality{perform}{\variousspeciality}}\\
	\capital{\skillref{rhetoric}}\\
	\capital{\skillref{subterfuge}}\\
	\bottomrule
}

For \skillref{craft}, \skillref{lore}, \skillref{perform}, many different versions of each skill exist.
The skill descriptions below list a number of possible versions of each skill, but a player may select any version they can imagine with the GM's permission.
A student may have a different rank for each version of the skill.
For example, they may have \skillrank{6}{3}{2} for \skillrefspeciality{craft}{woodcarving}, \skillrank{4}{4}{4} for \skillrefspeciality{craft}{glassblowing}, and \skillrank{0}{0}{0} for all other versions of the skill.
Each class or extracurricular that advances these skills will specify a version of the skill, or may allow you to advance any one version of the skill of your choice.

\section{Advancing Skills}
\seclabel{advancing-skills}

There is more to skill than simply level of education; there are several facets in which a student may develop each skill.
Each of a student wizard's skills is represented by a set of three numbers ranging from 0 to 7, such as \skillrank{6}{4}{3}.
The first number represents the depth of their education or training in a skill, and hence determines the maximum they are able to achieve.
The latter two numbers represent the breadth of their education, or how well practised they are.
A student lacking this breadth may have the potential to do very well, but their performance will be less reliable.

All students enter the school with their skills ranked \skillrank{0}{0}{0}.
They still roll 3 dice for {\tests}, so they're not totally incompetent.
But they've all just come out of primary education; they haven't yet specialised.
Characters advance their skills by taking classes and participating in extracurricular activities.
Each class or extracurricular will advance one or more skills.

Each time within a year that a character advances a skill, they add 1 to one of the three numbers representing their skill, working from left to right.
As such, a character cannot benefit from advancing the same skill more than three times in a year.
Furthermore, none of the numbers in a character's skill may exceed their current year level.
At maximum, a character may graduate with a skill ranked \skillrank{7}{7}{7}.

As an example, consider the \skillref{athletics} skill of Dalton Robins over his first four years of school.
Dalton enters school with his \skillref{athletics} ranked at \skillrank{0}{0}{0}, the same as everyone else.
In first year he participates in enough sport to advance his \skillref{athletics} twice, raising the first two numbers of his skill, making it \skillrank{1}{1}{0}.
In second year he does a little less sport and advances \skillref{athletics} only once, raising only the first number of his skill, making it \skillrank{2}{1}{0}.
In third year he does no sport at all, leaving his skill at \skillrank{2}{1}{0}.
In fourth year, feeling himself growing unfit, Dalton returns to sport with a vengeance and advances \skillref{athletics} four times.
The first three advances raise all three numbers of his skill, making it \skillrank{3}{2}{1}, but the fourth advance is wasted.

\section{Mundane Skills}
\seclabel{mundane-skills}

Mundane skills are those used in mundane actions---actions that would be possible even for those without magical talent or training.
They are rarely taught in classes at magical schools, and most students will develop them primarily through their extracurricular activities.

\skill{Aim}{aim}

Used to hit distant targets with something, be it thrown by hand, hit by a bat, or loosed from a bow.
Spells are usually targeted using the appropriate {\magicskill}, but some classes may teach students to aim them using their mundane ability. %TODO: Write this class and provide the relevant link.

\skill{Athletics}{athletics}

Used to run, jump, climb, swim, hike, ski, and generally to get about the place.
While many sports will require skill in \skillref{aim} or \skillref{fight}, most sports depend heavily, even primarily, on \skillref{athletics}.

\skill{Beast Handling}{beast-handling}

Used to understand beasts and interact with them: to calm them, tame them, ride them, train them, and command them.
While normally applied to animals, magical or mundane, some magical plants and fungi have sufficient will and motility that this skill can be useful in dealing with them.

\skill{Charm}{charm}

Used to make people like you: to befriend them, to build rapport with them, or to disarm them with a smile.
While \skillref{rhetoric} is used to make a persuasive argument, people who like you enough will often take your suggestions without any argument required.

\skill{Craft}{craft}

Used to make things with your hands.
Each \skillref{craft} skill is learned separately, and represents the ability to practise a particular craft.
Available crafts include the following:
\begin{itemize}
	\item Blacksmithing
	\item Carpentry
	\item Cooking
	\item Glassblowing
	\item Jewellery
	\item Masonry
	\item Pottery
	\item Seamstressing
	\item Woodcarving
	%TODO: Evaluate and maybe expand.
\end{itemize}

\skill{Fight}{fight}

Used to engage in hand-to-hand combat, either with weapons or without.
This involves both harming your opponent and preventing them from harming you.

\skill{Investigation}{investigation}

Used to uncover things by extended investigative effort.
This includes rifling a desk drawer for a document, ransacking a room for equipment, searching a library for pertinent information, or even unearthing a piece of juicy gossip.

\skill{Leadership}{leadership}

Used to command obedience from a position of authority or superiority.
This includes leading subordinates, whether they follow you willingly or due to your institutionally invested authority.
It also includes intimidating those who are physically or socially inferior to you.

\skill{Lore}{lore}

Used to know and recall information.
Each \skillref{lore} skill is learned separately, and represents knowledge of a different field.
Available fields of knowledge include the following (knowledge of magic is covered by {\magicskills}):
\begin{itemize}
	\item Botany
	\item Geography
	\item History
	\item Medicine
	\item Mycology %TODO: Roll this into Botany?
	\item Philosophy
	\item Politics
	\item Religion
	\item Zoology
	%TODO: Evaluate and maybe expand.
\end{itemize}

\skill{Observe}{observe}

Used to notice things, by vision, hearing, or even smell.
This includes spotting something out of place, such as spotting a piece of lab equipment out of place, or hearing someone sneaking up on you.
It also includes watching people carefully, such that you might pick up on signs they're hiding something or lying.

\skill{Perform}{perform}

Used to entertain people by putting on a show.
Each \skillref{lore} skill is learned separately, and represents a different form of entertainment.
Available methods of performance include the following:
\begin{itemize}
	\item Acting
	\item Dancing
	\item Drums
	\item Flute
	\item Singing
	\item Trumpet
	\item Violin
	%TODO: Evaluate and maybe expand.
\end{itemize}

\skill{Rhetoric}{rhetoric}

Used to persuade people with reason, logic, debate, and eloquence.
While \skillref{charm} can make someone like you, \skillref{rhetoric} can make them agree with you, even despite how they might feel about you personally.
\capital{\skillref{rhetoric}} is also used for most writing, a useful skill for a student.

\skill{Subterfuge}{subterfuge}

Used to conceal your covert activities.
This includes sneaking, hiding, stealing, smuggling, and even lying.

\section{Magic Skills}
\seclabel{magic-skills}

Magical skills represent a student's mastery of the various magical disciplines---they are primarily developed by attending classes.
They are primarily used in the actual practice of magic, though each doubles as a \skillref{lore} skill concerning magic of the relevant discipline.

\skill{Divination}{divination}

Used to obtain information by magic.
\capital{\skillref{divination}} allows a wizard to scry on remote locations, sense things beyond normal perception, pierce illusions, and otherwise to discover whatever they cannot learn mundanely.

\skill{Elementalism}{elementalism}

Used to manipulate the four classical elements: water, earth, fire, and air.
While not as broad in target as \skillref{kinetics}, \skillref{elementalism} is far more versatile in what it can do with the elements it affects.

\skill{Enchanting}{enchanting}

Used to bind magical power into objects.
\capital{\skillref{enchanting}} can invest objects with the ability to generate magical effects independently of a wizard, among other things.
It is also used to create casting tools and potions.

\skill{Illusion}{illusion}

Used to fool the senses by tricks of light, sound, scent, or other phenomena.
\capital{\skillref{illusion}} can efficiently create large or intricate effects, with the caveat that they have no physical substance.
Used to deceive or mislead, however, they can be very effective.

\skill{Kinetics}{kinetics}

Used to manipulate motion and forces.
In its most obvious form, this allows telekinesis.
However, it can also be applied to allow flight, warding, teleportation, and more.

\skill{Mentalism}{mentalism}

Used to manipulate the minds of people and beasts.
This includes putting them to sleep, subtly manipulating their thoughts, altering their memories, or even controlling their minds outright.
Applied to people, this is widely considered to be one of the more ethically troublesome forms of magic.

\skill{Metamagic}{metamagic}

Used to manipulate magic itself.
\capital{\skillref{metamagic}} spells often affect other spells, empowering them, cancelling them, or otherwise altering them.
They may also affect mana itself, generating it, or funnelling more of it into a single effect.

\skill{Necromancy}{necromancy}

Used to manipulate the dead, or to bring about death in the living.
In the former application, \skillref{necromancy} allows the animation of undead or communion with spirits.
In the latter application, it directly weakens, wounds, or kills, without any need for the fire of \skillref{elementalism} or the bludgeon of \skillref{kinetics}.
\capital{\skillref{necromancy}} is perhaps the most feared form of magic, and its use is often highly regulated.

\skill{Restoration}{restoration}

Used to repair damage, restoring things to their original state.
\capital{\skillref{restoration}} is most well known for its use upon people: healing.
However, it can also be used to heal animals and plants, and even to repair damaged or destroyed objects.

\skill{Transmutation}{transmutation}

Used to transform things into other things.
This may affect nonliving things, as in turning lead into gold; living things, as in turning people into animals; or turn one into the other, as in turning a person to stone.
